Title
========================================================

Data Preprocessing
-------------------------

### Retrieve the data
```{r}
setwd("~/Documents/Sandy/DA_HW1/code")
loansData <- read.csv("../data/loansData.csv")
```

### Data Cleaning
####Convert columns to numeric class, when applicable.
```{r}
loansData$Interest.Rate <- as.numeric(sub("%", "", loansData$Interest.Rate))
loansData$Debt.To.Income.Ratio <- as.numeric(sub("%", "", loansData$Debt.To.Income.Ratio))
loansData$Amount.Requested <- as.numeric(loansData$Amount.Requested)
loansData$Amount.Funded.By.Investors <- as.numeric(loansData$Amount.Funded.By.Investors)
loansData$Monthly.Income <- as.numeric(loansData$Monthly.Income)
loansData$Open.CREDIT.Lines <- as.numeric(loansData$Open.CREDIT.Lines)
loansData$Revolving.CREDIT.Balance <- as.numeric(loansData$Revolving.CREDIT.Balance)
loansData$Inquiries.in.the.Last.6.Months<- as.numeric(loansData$Inquiries.in.the.Last.6.Months)
loansData$Monthly.Income.log10 <- log10(as.numeric(loansData$Monthly.Income))
```

####Remove rows containing NAs.
```{r}
table(is.na(loansData))
loansData <- loansData[complete.cases(loansData),]
nrow(loansData) ## removed 2 cases
```

#### split FICO Range into min/max
```{r}
loansData$FICO.Range.min <- as.numeric(gsub("-(.*)","",loansData$FICO.Range))
loansData$FICO.Range.max <- as.numeric(gsub("(.*)-","",loansData$FICO.Range))
loansData$FICO.Range.mid <- (loansData$FICO.Range.max+loansData$FICO.Range.min)/2
```
#### convert loan length into numeric
```{r}
loansData$Loan.Length.months <- as.numeric(gsub(" months","",loansData$Loan.Length))
```
#### convert Employment Length to numeric
Assign 0 to <1 year, 10 to 10+ years
```{r}
loansData$Employment.Length.num <- gsub("< 1 year","0",loansData$Employment.Length)
loansData$Employment.Length.num <- gsub(" year(.*)","",loansData$Employment.Length.num)
loansData$Employment.Length.num <- gsub("n/a","",loansData$Employment.Length.num)
loansData$Employment.Length.num <- as.numeric(gsub("\\+","",loansData$Employment.Length.num))
```
#### convert some factor columns into numeric
```{r}
loansData$State.num <- as.numeric(loansData$State)
loansData$Loan.Purpose.num <- as.numeric(loansData$Loan.Purpose)
loansData$Home.Ownership.num <- as.numeric(loansData$Home.Ownership)
```
#### Check loansData table

```{r}
head(loansData)
names(loansData)
table(is.na(loansData)) # 77 NAs
summary(loansData)
```
 requested amounts are between 1000-35000. 

WHy is there a -0.01 min value for amount funded?
Interest rates range from 5.42 to 24.89
2 loan lengths: 36 months, 60 months
Debt to Income Ratio ranges from 0 to 34.91
FICO ranges from 640 to 834

#### Find missing values, check ranges
table(complete.cases(loansData))

Exploratory Analysis
-------------------------
### Check Interest rate vs FICO
```{r}
plot(jitter(loansData$FICO.Range.mid),
     jitter(loansData$Interest.Rate), col="blue", pch=19, cex=0.5)
smoothScatter((loansData$FICO.Range.min+loansData$FICO.Range.max)/2,loansData$Interest.Rate)
# plot shows decrease in interest rates when FICO score increases
boxplot(Interest.Rate ~ FICO.Range, data=loansData)
```
### try alternate colorings based on Loan.purpose, home.ownership, etc.
see if anything significant about home ownership
```{r}
plot(jitter((loansData$FICO.Range.min+loansData$FICO.Range.max)/2),
     jitter(loansData$Interest.Rate), col=as.numeric(loansData$Home.Ownership), pch=19, cex=0.4)
legend("topright", legend=unique(loansData$Home.Ownership), 
       col=unique(as.numeric(loansData$Home.Ownership)), pch=19)
plot(density(loansData$FICO.Range.mid), lwd=3, col="red")
```

#### Check FICO-related parameters
FICO scores are calculated using revolving credit balance, inquiries in the last 12 months (not 6),
open credit lines
```{r}
plot(jitter(loansData$FICO.Range.min),jitter(loansData$Inquiries.in.the.Last.6.Months), col="blue", pch=19, cex=0.5)
```

### Loan Length
As of this writing, Lending Club only allows 36-month terms for loans of $1000-$15,975,
so naturally larger loan lengths will tend to be for higher amounts.
```{r}
plot(jitter(loansData$FICO.Range.min),jitter(loansData$Amount.Requested), col=as.numeric(loansData$Loan.Length), pch=19, cex=0.5)
legend("topright", legend=unique(loansData$Loan.Length), col=unique(as.numeric(loansData$Loan.Length)), pch=19, cex=0.9)
plot(jitter(loansData$Amount.Requested), jitter(loansData$Interest.Rate), col=loansData$Employment.Length.num, pch=19, cex=0.5)
plot(jitter(loansData$FICO.Range.min), jitter(loansData$Debt.To.Income.Ratio), col="blue", pch=19, cex=0.5)
```
Slight relationship --> higher FICO scores, more likely to have higher Debt-to-income ratio

### Correlations
#### Check Interest Rates
```{r}
loansDataCor <- cor(loansData[,c(1,2,3,19,22,6,21,23,9,18,11,12,13)])
loansDataCor
```
Correlation indicates the following vars are highly correlated with Interest Rate:
FICO.Range.mid(-0.7092), Loan.Length.months(0.4235), Amount.Funded.By.Investors(0.336),
Amount.Requested(0.331)
As expected, state does not correlate well with FICO score (not in FICO score)
Best vars are
Debt.To.Income.Ratio(-0.2169), Home.Ownership.num (-0.156), Monthly.Income(0.122)
Oddly, Inquiries in the last 6 months is low

Modeling
-------------------------
### Start with basic linear regression model, using FICO score only
```{r}
palette(rainbow(14))
loansDataLm1 <- lm(Interest.Rate ~ FICO.Range.mid, data=loansData)
summary(loansDataLm1)
par(mfrow=c(1,2))
plot(loansData$Interest.Rate, loansDataLm1$residuals, col=as.numeric(loansData$Loan.Purpose), pch=19, cex=0.5)
legend("bottomright",legend=unique(loansData$Loan.Purpose),col=as.numeric(unique(loansData$Loan.Purpose)),pch=19, cex=0.6)
plot(loansDataLm1$fitted.values, loansDataLm1$residuals, col=as.numeric(loansData$Loan.Purpose), pch=19, cex=0.5)
```
### add in Loan Lengths
should probably make it into the factor version
```{r}
loansDataLm2 <- lm(Interest.Rate ~ FICO.Range.mid + Amount.Requested + as.factor(Loan.Length), data=loansData)
summary(loansDataLm2)
par(mfrow=c(1,2))
plot(loansData$Interest.Rate, loansDataLm2$residuals, col=as.numeric(loansData$Loan.Purpose), pch=19, cex=0.5)
plot(loansDataLm2$fitted.values, loansDataLm2$residuals, col=as.numeric(loansData$Loan.Purpose), pch=19, cex=0.5)
```

### add in Inquiries instead
Slightly better, R changed by 0.01
```{r}
loansDataLm3 <- lm(Interest.Rate ~ FICO.Range.mid + Amount.Requested + as.factor(Loan.Length)  + Inquiries.in.the.Last.6.Months, data=loansData)
summary(loansDataLm3)
par(mfrow=c(1,2))
plot(loansData$Interest.Rate, loansDataLm3$residuals, col=as.numeric(loansData$Loan.Purpose), pch=19, cex=0.5)
plot(loansDataLm3$fitted.values, loansDataLm3$residuals, col=as.numeric(loansData$Loan.Purpose), pch=19, cex=0.5)
par(mfrow=c(1,1))
plot(loansDataLm3,which=2)
confint(loansDataLm3)
```

### Check confounding
```{r}
confoundlm <- lm(Amount.Requested ~ as.factor(Loan.Length), data=loansData)
summary(confoundlm)
```

### Retry using FICO range as factor
with confounder, adjusted R-squared: of 0.796
```{r}
loansDataLm4 <- lm(Interest.Rate ~ as.factor(FICO.Range) + Amount.Requested + as.factor(Loan.Length) + Inquiries.in.the.Last.6.Months, data=loansData)
summary(loansDataLm4)
par(mfrow=c(1,2))
plot(loansData$Interest.Rate, loansDataLm4$residuals, col=as.numeric(loansData$Loan.Purpose), pch=19, cex=0.5)
plot(loansDataLm4$fitted.values, loansDataLm4$residuals, col=as.numeric(loansData$Loan.Purpose), pch=19, cex=0.5)
# P value is >0.001 but <0.01, so not sure if OK
par(mfrow=c(1,1))
plot(loansDataLm4,which=2)
```

Final Plots
------------
